\textbf{ Interpretable Gaussian Processes for Stellar Variability }

### 2021-01-14 tex/ms.tex
more text

we may write the mean and covariance of our flux GP as

@@

are the mean and covariance of the GP in the spherical harmonics basis.

@@

\subsection{Marginalizing over inclination}
\label{sec:inclination}
%
As we mentioned in the previous section,
the equations for the mean and covariance of the flux GP
(Equations~\ref{eq:mean_f} and \ref{eq:cov_f}, respectively) are conditioned
on specific values of the stellar and spot properties. To obtain the posterior distribution
for these parameters, we must typically resort to numerical sampling techniques,
which often scale steeply with the number of parameters. It is therefore generally
desirable to keep the total number of parameters small, especially when
employing the GP in an ensemble setting.
In such a setting, we might have light curves from $M$ stars, all of which
we believe to have similar spot properties (perhaps because they have
similar spectral types and rotation periods, for example).
The total number of parameters in our problem is therefore
%
\begin{align}
    N = 4 M + 5
    \quad,
\end{align}
%
since each of the stars will have their own set of 4 stellar properties
(an inclination, a period, and usually two limb darkening coefficients)
but will all share the same 5 spot properties
$\pmb{\theta}_\bullet$ (by assumption).
For a reasonably sized ensemble of $M=100$
stars, we would have to sample over $N = 405$ parameters.
While large, this number is certainly not absurd, especially by modern standards.
However, it does pose
a problem when considering how complex the posterior distribution for the
spot mapping problem can be. In addition to strong nonlinear degeneracies
between some of the parameters (such as the contrast $c$ and the
number of spots $n$), the posterior is often multimodal, especially in the
stellar inclinations. While modern sampling schemes such as Hamiltonian
Monte Carlo and Nested Sampling may in principle be able to deal with these
issues, in practice it can be very difficult to obtain convergence in a
reasonable amount of time.

@@

\subsection{Small spots}
\label{sec:tinyspots}
\xxx{TODO: The model is limited in resolution to features with radii $r \gtrsim 10^\circ$.
    Typical sun spots are MUCH smaller than that. However, our model can still
    correctly infer latitudinal parameters even when spots are really tiny
    (i.e., no bias due to the limited resolution). We might try to improve the
    numerical stability of the model at higher spherical harmonic degree, but
    that's a task for a future paper.}

@@

\subsection{Application to transits and RV}
\label{sec:other-work}
\xxx{TODO: The covariance in the spherical harmonic basis can be used as a prior
    in many other settings.}

@@

\subsection{Caveats}
\label{sec:caveats}
\xxx{TODO: Miscellaneous caveats.}
\xxx{
    \begin{enumerate}
        \item The latitude distribution is not Gaussian, so careful
              when interpreting $\mu_\phi$ and $\sigma_\phi$
        \item The number of spots does not have to be an integer. This is
              fine, since it makes the model more flexible.
        \item Careful with light curves with large amplitudes, although
              Figure~\ref{fig:calibration_hicontrast}
              has max amplitudes of more than 10\% and works great.
        \item No differential rotation / spot evolution yet! We're working on it.
        \item We might want a shrinkage prior on sigma, since it's hard to
              break the polar spot degeneracy with single-band light curves.
    \end{enumerate}
}

@@

@@

          or even radial velocity searches for exoplanets.

@@

The next papers in this series will focus on (II) a more rigorous
treatment of time variability and (III) explicit marginalization over
the remaining stellar parameters. In particular, paper II will focus
on modeling differential rotation, whose effect on the covariance of the
process can be derived in a similar fashion to what we did here. This
will enable direct inference about the differential rotation rates and
spot evolution timescales of stars, processes whose effects on light curves
are too similar for current methodology to reliably discern between them.
Paper III will focus on marginalization over the stellar rotation period
and limb darkening coefficients, which can also be done analytically
under certain choices of prior. This will eliminate all per-star
hyperparameters in the expression for the GP covariance, greatly
speeding up inference for large ensembles of stellar light curves.

@@

Our goal in this section is to find closed-form solutions to the
first and second moments of the spherical harmonic representation of the
stellar surface $\mathbbb{y}$,
%
\begin{align}
    \label{eq:exp_y_app}
    \mathrm{E} \Big[ \mathbbb{y} \, \Big| \, \pmb{\theta}_\bullet \Big]
     & =
    \int \mathbbb{y}(\mathbbb{x} ) \, p(\mathbbb{x} \, \big| \, \pmb{\theta}_\bullet)\mathrm{d}\mathbbb{x}
    \\
    \label{eq:exp_yy_app}
    \mathrm{E} \Big[ \mathbbb{y} \, \mathbbb{y}^\top \, \Big| \, \pmb{\theta}_\bullet \Big]
     & =
    \int \mathbbb{y}(\mathbbb{x} ) \mathbbb{y}^\top(\mathbbb{x} ) \, p(\mathbbb{x} \, \big| \, \pmb{\theta}_\bullet)\mathrm{d}\mathbbb{x}
    \quad,
\end{align}
%
which are linearly related to the mean and the covariance of our GP in flux
(\S\ref{sec:gp}).
%
Recall that $\mathbbb{x}$ is a vector of parameters describing the exact
configuration of features on the surface of a star, and
$p(\mathbbb{x} \, \big| \, \pmb{\theta}_\bullet)$ is its probability
density function conditioned on hyperparameters $\pmb{\theta}_\bullet$,
which describe the \emph{distribution} of the features on the surface
of one or many stars.
%
As we are specifically interested in modeling the effect of star spots
on stellar light curves, we let
%
\begin{align}
    \label{eq:x}
    \mathbbb{x} =
    \left(
    \mathbb{n} \,\,\,\,
    \mathbb{c}_0 \,\, \cdots \,\, \mathbb{c}_{\mathbb{n}-1} \,\,\,\,
    \bblambda_0 \,\, \cdots \,\, \bblambda_{\mathbb{n}-1} \,\,\,\,
    \bbphi_0 \,\, \cdots \,\, \bbphi_{\mathbb{n}-1} \,\,\,\,
    \mathbb{r}_0 \,\, \cdots \,\, \mathbb{r}_{\mathbb{n}-1}
    \right)^\top
\end{align}
%
and
%
\begin{align}
    \label{eq:RRs}
    \mathbbb{y}(\mathbbb{x}) =
    \sum_{i=0}^{\mathbb{n}-1}
    \mathbb{c}_i
    \,
    \mathbf{R}_{\hat{\mathbf{y}}}(\bblambda_i)
    \,
    \mathbf{R}_{\hat{\mathbf{x}}}(\bbphi_i)
    \,
    \mathbf{s}(\mathbb{r}_i)
    \quad,
\end{align}
%
where $\mathbb{n}$ is the total number of spots,
$\mathbb{c}_i$ is the contrast of the $i^\mathrm{th}$ spot,
$\bblambda_i$ is its longitude, $\bbphi_i$ is its latitude,
and $\mathbb{r}_i$ is its radius.
The vector function $\mathbf{s}(\mathbb{r}_i)$
returns the spherical harmonic expansion of a negative unit brightness
circular spot of radius $\mathbb{r}_i$ at $\bblambda = \bbphi = 0$,
$\mathbf{R}_{\hat{\mathbf{x}}}(\bbphi_i)$ is the Wigner matrix that rotates the
expansion about $\hat{\mathbf{x}}$ such that the spot is centered at a
latitude $\bbphi_i$, and $\mathbf{R}_{\hat{\mathbf{y}}}(\bblambda_i)$ is the Wigner
matrix that then rotates the
expansion about $\hat{\mathbf{y}}$ such that the spot is centered at a
longitude $\bblambda_i$; these three functions are detailed in the sections below.
%
Equation~(\ref{eq:RRs}) thus provides a way of converting a random variable
$\mathbbb{x}$ describing the size, brightness, and position of spots to the
corresponding representation in terms of spherical harmonics.
%
Regarding this equation,
two things should be noted. First, we define $\mathbbb{y}$ relative to
a baseline of zero: i.e., a star with no spots on it will have
$\mathbf{y} = \mathbf{0}$. Second, and more importantly,
we are not interested in any specific value of
$\mathbbb{y}$; rather, we would like to know its expectation value under
the probability distribution governing the different spot properties $\mathbbb{x}$,
i.e., $p(\mathbbb{x} \, \big| \, \pmb{\theta}_\bullet)$.
%

@@

is the first moment of the GP in the spherical harmonic basis

@@

\subsection{Second moment}
\label{sec:inc-mom2}
%
The expression for the second moment is
%
\begin{align}
    \mathbf{E}_I
     & \equiv
    \int
    \pmb{\mathcal{A}}(\mathbb{I}, P, \mathbf{u}) \,
    \mathbf{E}_y \,
    \pmb{\mathcal{A}}^\top(I, P, \mathbf{u}) \,
    p(\mathbb{I}) \,
    \mathrm{d}\mathbb{I}
\end{align}
%
where
%
\begin{align}
    \mathbf{E}_y
     & \equiv
    \mathrm{E} \Big[ \mathbbb{y} \, \mathbbb{y}^\top \, \Big| \, \pmb{\theta}_\bullet \Big]
\end{align}
%
is the second moment of the GP in the spherical harmonic basis
(Equation~\ref{eq:exp_yy}). We can use Equations~(\ref{eq:Arows})
and (\ref{eq:akT})
to express the element at index $k, k'$ (corresponding to the covariance of the
GP between times $t = t_k$ and $t' = t_{k'}$) as
%
\begin{align}
    \label{eq:Eikkp}
    \left(E_I\right)_{k, k'}
     & =
    \int
    \mathbf{a}_k^\top(\mathbb{I})
    \mathbf{E}_y \,
    \mathbf{a}_{k'}(\mathbb{I})
    p(\mathbb{I}) \,
    \mathrm{d}\mathbb{I}
    \nonumber \\
     & =
    \mathbf{r}^\top \,
    \mathbf{A_1} \,
    \left(\mathbf{E}_{y''}\right)_{k,k'}
    \mathbf{A_1}^\top \,
    \mathbf{r}
    \quad,
\end{align}
%
where
%
\begin{align}
    \label{eq:Eyppkkp}
    \left(\mathbf{E}_{y''}\right)_{k,k'}
     & =
    \int
    \mathbf{R}_{\hat{\mathbf{x}}}\left(-\mathbb{I}\right) \,
    \left(\mathbf{E}_{y'}\right)_{k,k'} \,
    \mathbf{R}_{\hat{\mathbf{x}}}^\top\left(-\mathbb{I}\right) \,
    p(\mathbb{I}) \,
    \mathrm{d}\mathbb{I}
\end{align}
%
is the expectation of $\mathbbb{y}\,\mathbbb{y}^\top$ in the
observer's frame at times $t = t_k$ and $t' = t_{k'}$
and
%
\begin{align}
    \left(\mathbf{E}_{y'}\right)_{k,k'}
     & \equiv
    \mathbf{R}_{\hat{\mathbf{z}}}\left(\frac{2\pi}{P}t_k\right) \,
    \mathbf{R}_{\hat{\mathbf{x}}}\left(\frac{\pi}{2}\right) \,
    \mathbf{E}_y \,
    \mathbf{R}_{\hat{\mathbf{x}}}^\top\left(\frac{\pi}{2}\right) \,
    \mathbf{R}_{\hat{\mathbf{z}}}^\top\left(\frac{2\pi}{P}t_{k'}\right)
\end{align}
%
is the expectation of $\mathbbb{y}\,\mathbbb{y}^\top$ in the
polar frame at times $t = t_k$ and $t' = t_{k'}$.
%
The rest of the computation follows what we did in \S\ref{sec:lat},
except that the number of operations required to compute $\mathbf{E}_I$
is a factor of $K^2$ larger than in the computation of expectations
like $\mathbf{E}_\phi$ (Equation~\ref{eq:Ephi}). That is because we
must compute the integral of all terms of a matrix
for \emph{each} of the $K^2$ elements of $\mathbf{E}_I$.
We discuss in \S\ref{sec:inc-speedup} below strategies that can drastically
improve the computational scaling of marginalizing over the inclination.

>>>>

we may write the mean and covariance of our GP as

@@

are the mean and covariance of the distribution over spherical harmonic coefficients.

@@

\subsection{Marginalizing over inclination}
\label{sec:inclination}
%
As we mentioned in the previous section,
the equations for the mean and covariance of our GP
(Equations~\ref{eq:mean_f} and \ref{eq:cov_f}, respectively) are conditioned
on specific values of the stellar and spot properties. To obtain the posterior distribution
for these parameters, we must typically resort to numerical sampling techniques,
which often scale steeply with the number of parameters. It is therefore generally
desirable to keep the total number of parameters small, especially when
employing the GP in an ensemble setting.
In such a setting, we might have light curves from $M$ stars, all of which
we believe to have similar spot properties (perhaps because they have
similar spectral types and rotation periods, for example).
The total number of parameters in our problem is therefore
%
\begin{align}
    N = 4 M + 5
    \quad,
\end{align}
%
since each of the stars will have their own set of 4 stellar properties
(an inclination, a period, and usually two limb darkening coefficients)
but will all share the same 5 spot properties
$\pmb{\theta}_\bullet$ (by assumption).
For a reasonably sized ensemble of $M=100$
stars, we would have to sample over $N = 405$ parameters.
While large, this number is certainly not absurd, especially by modern standards.
However, it does pose
a problem when considering how complex the posterior distribution for the
spot mapping problem can be. In addition to strong nonlinear degeneracies
between some of the parameters (such as the contrast $c$ and the
number of spots $n$), the posterior is often multimodal, especially in the
stellar inclinations. While modern sampling schemes such as Hamiltonian
Monte Carlo and Nested Sampling may in principle be able to deal with these
issues, in practice it can be very difficult to obtain convergence in a
reasonable amount of time.

@@

\subsection{Small spots}
\label{sec:tinyspots}

One of the biggest downsides of adopting a spherical harmonic representation
of the stellar surface as the foundation of our GP is the inherent limitation
it imposes on the resolution of surface features. In order to maximize
computational efficiency and numerical stability, our default approach is to
model the surface using an expansion up to degree $l_\mathrm{max} = 15$,
which can model features only as small as
$\sim \nicefrac{180^\circ}{15} = 12^\circ$ on the surface. Even on scales
slightly larger than this, the presence of ringing can be seen
(c.f. panel (b) in Figure~\ref{fig:samples}, where ringing can just barely be
seen in the equatorial region of the maps). The spherical harmonic basis consists
of global modes, all of which contribute to the intensity everywhere on the surface.
Localized features require constructive interference of modes inside
and destructive interference of modes outside, often leading to a wave-like ringing
pattern that gets worse as the size of the features gets smaller. Taken at
face value, this might suggest that a different basis---such as the common
choice of pixels on a grid, or perhaps a localized wavelet basis---would be better
at modeling small spots. While this is probably true, a different choice of
basis would not admit the same analytic solutions to the integrals needed to
compute the GP covariance. And it is precisely the ability to compute these
integrals analytically (Appendix~\ref{sec:integrals}) that makes the evaluation
of the marginal likelihood tractable; it is difficult to envision an approach
that uses a \emph{local} basis and simultaneously enjoys the same
analyticity under integration.

In principle, small spots can be modeled with negligible ringing by simply
increasing the degree of the spherical harmonic expansion. As we discuss in
\S\ref{sec:implementation}, however, the algorithm presented here becomes
unstable for $l \gtrsim 15$, so doing so would require a reparametrization
of the equations in the Appendix. However, as we showed in
\S\ref{sec:calibration}, the current implementation of the GP is suitable
to modeling light curves of stars with spots smaller than the limiting
resolution of $\sim 10^\circ$. Consider Figure~\ref{fig:calibration_tinyspots},
which shows the results of doing inference on an ensemble of light curves
of stars with small ($r = 3^\circ$) spots; these are comparable to some of
the larger spots seen on the Sun with diameters of about $70,000$ km.
Even though our inferred radius and contrast are wrong, the fact that the radius
posterior peaks
at the lower prior bound of $10^\circ$ is strongly suggestive of the presence
of spots smaller than the resolution of the model. Moreover, the
latitude mode and standard deviation posteriors are \emph{unbiased}, and we
correctly infer the presence of low-variance, mid-latitude spots.
With the above caveats in mind, our GP can therefore still be used to model
stars with small spots.

@@

@@

\subsection{Caveats}
\label{sec:caveats}

We conclude our discussion with a list of several caveats that should be kept
in mind when using our algorithm and/or its \Python implementation.

\xxx{TODO}

\begin{enumerate}
    \item \emph{The assumed latitude distribution is not Gaussian}.
          Careful when interpreting $\mu_\phi$ and $\sigma_\phi$.
    \item \emph{The number of spots $n$ does not have to be an integer}.
          This is fine, since it makes the model more flexible.
    \item \emph{Care should be taken when modeling large-amplitude light curves}.
          Careful with light curves with large amplitudes, although
          Figure~\ref{fig:calibration_hicontrast}
          has max amplitudes of more than 10\% and works great.
    \item \emph{Keep in mind the polar spot degeneracy}.
          We might want a shrinkage prior on sigma, since it's hard to
          break the polar spot degeneracy with single-band light curves.
    \item \emph{Careful with the data}. Jitter term, baseline, etc.
    \item \emph{Careful with sample selection}. Even non-variable light curves.
          If you're worried, use a mixture model.
\end{enumerate}

@@

\subsection{Modeling transits and radial velocity datasets}
\label{sec:transits-rvs}

Finally, we would like to note that the GP developed in this paper is not
limited to modeling (rotational) light curves of stars. We derived the
covariance structure of spotted stellar surfaces in the spherical harmonic
basis, which we then linearly transformed into flux space to obtain the
light curve GP presented above. However, our GP may be used to model
\emph{any} kind of observation that
is linearly related to the spherical harmonic representation. If $\mathbf{A}$
is the linear operator that transforms the spherical harmonic representation
$\mathbf{y}$ to the data vector $\mathbf{d}$ via
$\mathbf{d} = \mathbf{A} \mathbf{y}$, then the mean and covariance of the
GP model for $\mathbf{d}$ are given by
%
\begin{align}
    \pmb{\mu}_\mathbf{d}    & = \mathbf{A} \, \pmb{\mu}_\mathbf{y}
    \\
    \pmb{\Sigma}_\mathbf{d} & = \mathbf{A} \, \pmb{\Sigma}_\mathbf{y} \, \mathbf{A}^\top
    \quad,
\end{align}
%
respectively, where $\pmb{\mu}_\mathbf{y}$ and $\pmb{\Sigma}_\mathbf{y}$
are the mean and covariance in the spherical harmonic basis derived in this paper.
In \citet{Luger2019} we showed that
occultation light curves are also linearly related to the
spherical harmonic representation of the stellar surface, so it is
straightforward to use our GP to model transits of planets across spotted
stars, either to marginalize over the spot variability or to constrain the
surface map of the star. In this context, $\mathbf{A}$ may be computed via the
\textsf{design\_matrix()} method of a \textsf{Map} instance of the \starry
package.

The formalism developed here may also be extended to model radial
velocity (RV) datasets, although this requires a bit more work. The
instantaneous radial
velocity shift $v$ induced by a rotating spotted star may be approximated as
%
\begin{align}
    v & = \frac{\iint_S I \, V \, \mathrm{d} S}{\iint_S I \mathrm{d} S}
\end{align}
%
where $I$ is the stellar intensity at a point on the surface, $V$ is the
radial component of the rotational velocity vector at that point, and the
integral is taken over the projected disk of the star.
%
If we expand the surface intensity distribution in spherical harmonics,
the integral in the denominator is just a classical \starry integral, as it
is just the disk-integrated intensity (i.e., the flux).
%
The numerator may also be computed following the \starry formalism,
provided we weight the surface intensity representation by the velocity
field $V$. In the case of rigid body rotation, $V$ is exactly a dipole ($l = 1$)
field, so the quantity $I V$ can be expressed exactly as a product of
spherical harmonics.%
\footnote{
    Higher order effects such as differential rotation and convective
    blueshift can be easily modeled with a higher degree expansion of $u$.
}
Since spherical harmonics are closed under multiplication,
we may write $I V$ as a linear combination of spherical harmonics, meaning
the integral in the numerator is \emph{also} a \starry integral.
For some linear operators $\mathbf{A}$ and $\mathbf{B}$, we may therefore
write
%
\begin{align}
    \mathbbb{v} & = \frac{\mathbf{B} \, \mathbbb{y}}{\mathbf{A} \, \mathbbb{y}}
\end{align}
%
where $\mathbbb{v}$ is the random variable representing the observed radial
velocity timeseries and
$\mathbbb{y}$ is the Gaussian random variable representing the spherical harmonic
coefficients describing the stellar surface.
%
Since the $\mathbbb{v}$ is the ratio of two Gaussian random variables, its
distribution is not Gaussian. However, we can still compute the first two moments
of the distribution of $\mathbbb{v}$ to derive a Gaussian approximation to it
(similar to what we did in \S\ref{sec:baseline}), which will yield the mean
and covariance of the Gaussian process representation of the radial velocity
timeseries.
%
Given the complexity of the operations described above,
we defer this calculation (and the calibration of the resulting GP model)
to future work.

@@

          or even radial velocity searches for exoplanets. The latter will
          be the subject of a future paper in this series.

@@

The next papers in this series will focus on (in no particular order)
\begin{itemize}
    \item \emph{A more rigorous treatment of time variability}.
          This paper will focus
          on modeling differential rotation, whose effect on the covariance of the
          process can be derived in a similar fashion to what we did here. This
          will enable direct inference about the differential rotation rates and
          spot evolution timescales of stars, processes whose effects on light curves
          are too similar for current methodology to reliably discern between them.
    \item \emph{Explicit marginalization over the remaining stellar parameters}.
          These include the stellar rotation period
          and limb darkening coefficients, which can be marginalized over analytically
          under certain choices of prior. This will eliminate all per-star
          hyperparameters in the expression for the GP covariance, greatly
          speeding up inference for large ensembles of stellar light curves.
    \item \emph{Extension of this formalism to radial velocity datasets}.
          As we discussed in \S\ref{sec:transits-rvs}, it is possible to extend
          the methodology presented here to model the contribution of stellar
          surface variability to radial velocity measurements, which can be used
          (for instance) to mitigate systematics in extreme precision radial velocity
          (EPRV) searches for exoplanets.
    \item \emph{Extension of this formalism to Doppler imaging}.
          As we show in upcoming work, it is possible to derive an exact
          linear relationship between the wavelength-dependent spherical harmonic
          representation of a rotating star and its time-variable spectrum.
          This linearity makes it possible to adapt our GP formalism to the Doppler
          imaging problem, providing an efficient marginal likelihood function
          for rigorous inference studies.
\end{itemize}

@@

Our goal in this section is to find closed-form solutions to the
first and second moments of the spherical harmonic representation of the
stellar surface $\mathbbb{y}$,
%
\begin{align}
    \label{eq:exp_y_app}
    \mathrm{E} \Big[ \mathbbb{y} \, \Big| \, \pmb{\theta}_\bullet \Big]
     & =
    \int \mathbbb{y}(\mathbbb{x} ) \, p(\mathbbb{x} \, \big| \, \pmb{\theta}_\bullet)\mathrm{d}\mathbbb{x}
    \\
    \label{eq:exp_yy_app}
    \mathrm{E} \Big[ \mathbbb{y} \, \mathbbb{y}^\top \, \Big| \, \pmb{\theta}_\bullet \Big]
     & =
    \int \mathbbb{y}(\mathbbb{x} ) \mathbbb{y}^\top(\mathbbb{x} ) \, p(\mathbbb{x} \, \big| \, \pmb{\theta}_\bullet)\mathrm{d}\mathbbb{x}
    \quad,
\end{align}
%
which are linearly related to the mean and the covariance of our GP
(\S\ref{sec:gp}).
%
Recall that $\mathbbb{x}$ is a vector of parameters describing the exact
configuration of features on the surface of a star, and
$p(\mathbbb{x} \, \big| \, \pmb{\theta}_\bullet)$ is its probability
density function conditioned on hyperparameters $\pmb{\theta}_\bullet$,
which describe the \emph{distribution} of the features on the surface
of one or many stars.
%
As we are specifically interested in modeling the effect of star spots
on stellar light curves, we let
%
\begin{align}
    \label{eq:x}
    \mathbbb{x} =
    \left(
    \mathbb{n} \,\,\,\,
    \mathbb{c}_0 \,\, \cdots \,\, \mathbb{c}_{\mathbb{n}-1} \,\,\,\,
    \bblambda_0 \,\, \cdots \,\, \bblambda_{\mathbb{n}-1} \,\,\,\,
    \bbphi_0 \,\, \cdots \,\, \bbphi_{\mathbb{n}-1} \,\,\,\,
    \mathbb{r}_0 \,\, \cdots \,\, \mathbb{r}_{\mathbb{n}-1}
    \right)^\top
\end{align}
%
and
%
\begin{align}
    \label{eq:RRs}
    \mathbbb{y}(\mathbbb{x}) =
    \sum_{i=0}^{\mathbb{n}-1}
    \mathbb{c}_i
    \,
    \mathbf{R}_{\hat{\mathbf{y}}}(\bblambda_i)
    \,
    \mathbf{R}_{\hat{\mathbf{x}}}(\bbphi_i)
    \,
    \mathbf{s}(\mathbb{r}_i)
    \quad,
\end{align}
%
where $\mathbb{n}$ is the total number of spots,
$\mathbb{c}_i$ is the contrast of the $i^\mathrm{th}$ spot,
$\bblambda_i$ is its longitude, $\bbphi_i$ is its latitude,
and $\mathbb{r}_i$ is its radius.
The vector function $\mathbf{s}(\mathbb{r}_i)$
returns the spherical harmonic expansion of a negative unit brightness
circular spot of radius $\mathbb{r}_i$ at $\bblambda = \bbphi = 0$,
$\mathbf{R}_{\hat{\mathbf{x}}}(\bbphi_i)$ is the Wigner matrix that rotates the
expansion about $\hat{\mathbf{x}}$ such that the spot is centered at a
latitude $\bbphi_i$, and $\mathbf{R}_{\hat{\mathbf{y}}}(\bblambda_i)$ is the Wigner
matrix that then rotates the
expansion about $\hat{\mathbf{y}}$ such that the spot is centered at a
longitude $\bblambda_i$; these three functions are detailed in the sections below.
%
Equation~(\ref{eq:RRs}) thus provides a way of converting a random variable
$\mathbbb{x}$ describing the size, brightness, and position of spots to the
corresponding representation in terms of spherical harmonics.
%
Regarding this equation,
two things should be noted. First, we define $\mathbbb{y}$ relative to
a baseline of zero: i.e., a star with no spots on it will have
$\mathbf{y} = \mathbf{0}$. Second, and more importantly,
we are not interested in any specific value of
$\mathbbb{y}$; rather, we would like to know its expectation value under
the probability distribution governing the different spot properties $\mathbbb{x}$,
i.e., $p(\mathbbb{x} \, \big| \, \pmb{\theta}_\bullet)$.
%

@@

is the first moment of the distribution over spherical harmonic coefficients

@@

\subsection{Second moment}
\label{sec:inc-mom2}
%
The expression for the second moment is
%
\begin{align}
    \mathbf{E}_I
     & \equiv
    \int
    \pmb{\mathcal{A}}(\mathbb{I}, P, \mathbf{u}) \,
    \mathbf{E}_y \,
    \pmb{\mathcal{A}}^\top(I, P, \mathbf{u}) \,
    p(\mathbb{I}) \,
    \mathrm{d}\mathbb{I}
\end{align}
%
where
%
\begin{align}
    \mathbf{E}_y
     & \equiv
    \mathrm{E} \Big[ \mathbbb{y} \, \mathbbb{y}^\top \, \Big| \, \pmb{\theta}_\bullet \Big]
\end{align}
%
is the second moment of the distribution over spherical harmonic coefficients
(Equation~\ref{eq:exp_yy}). We can use Equations~(\ref{eq:Arows})
and (\ref{eq:akT})
to express the element at index $k, k'$ (corresponding to the covariance of the
GP between times $t = t_k$ and $t' = t_{k'}$) as
%
\begin{align}
    \label{eq:Eikkp}
    \left(E_I\right)_{k, k'}
     & =
    \int
    \mathbf{a}_k^\top(\mathbb{I})
    \mathbf{E}_y \,
    \mathbf{a}_{k'}(\mathbb{I})
    p(\mathbb{I}) \,
    \mathrm{d}\mathbb{I}
    \nonumber \\
     & =
    \mathbf{r}^\top \,
    \mathbf{A_1} \,
    \left(\mathbf{E}_{y''}\right)_{k,k'}
    \mathbf{A_1}^\top \,
    \mathbf{r}
    \quad,
\end{align}
%
where
%
\begin{align}
    \label{eq:Eyppkkp}
    \left(\mathbf{E}_{y''}\right)_{k,k'}
     & =
    \int
    \mathbf{R}_{\hat{\mathbf{x}}}\left(-\mathbb{I}\right) \,
    \left(\mathbf{E}_{y'}\right)_{k,k'} \,
    \mathbf{R}_{\hat{\mathbf{x}}}^\top\left(-\mathbb{I}\right) \,
    p(\mathbb{I}) \,
    \mathrm{d}\mathbb{I}
\end{align}
%
is the expectation of $\mathbbb{y}\,\mathbbb{y}^\top$ in the
observer's frame at times $t = t_k$ and $t' = t_{k'}$
and
%
\begin{align}
    \left(\mathbf{E}_{y'}\right)_{k,k'}
     & \equiv
    \mathbf{R}_{\hat{\mathbf{z}}}\left(\frac{2\pi}{P}t_k\right) \,
    \mathbf{R}_{\hat{\mathbf{x}}}\left(\frac{\pi}{2}\right) \,
    \mathbf{E}_y \,
    \mathbf{R}_{\hat{\mathbf{x}}}^\top\left(\frac{\pi}{2}\right) \,
    \mathbf{R}_{\hat{\mathbf{z}}}^\top\left(\frac{2\pi}{P}t_{k'}\right)
\end{align}
%
is the expectation of $\mathbbb{y}\,\mathbbb{y}^\top$ in the
polar frame at times $t = t_k$ and $t' = t_{k'}$.
%
The rest of the computation follows what we did in \S\ref{sec:lat},
except that the number of operations required to compute $\mathbf{E}_I$
is a factor of $K^2$ larger than in the computation of expectations
like $\mathbf{E}_\phi$ (Equation~\ref{eq:Ephi}). That is because we
must compute the integral of all terms of a matrix
for \emph{each} of the $K^2$ elements of $\mathbf{E}_I$.
We discuss in \S\ref{sec:inc-speedup} below strategies that can drastically
improve the computational scaling of marginalizing over the inclination.
